---
title: "5_results"
author: "Steve Harris"
date:	"30 Dec 2015"
output: md_document
---
```{r include=FALSE, echo=FALSE}
source("5_results_01start.R")
```

# Results

48 hospitals reported 20893 visits over 435 months. 2694 visits (12.9\%) did not meet the inclusion criteria including 1860 (8.9\%) repeat rather than first visits, and 586 (2.8\%) patients recently discharged from critical care. Data linkage was incomplete (\< 80\%) for 66 (15\%) study-months excluding a further 2440 (11.7\%) visits. Therefore 15759 patients were recruited to the study, of which 15158 (96.1\%) completed follow-up without error and were available for analysis (Figure 1). Final data linkage (ward visits to critical care admissions) was 93\% complete.

## Participating hospitals
```{r include=FALSE, echo=FALSE}
source("5_results_04hospitals.R")
```

The participating hospitals comprised `r n.teach` teaching and `r n.dgh` general hospitals that each collected data between September 2010 and December 2011 for a median of `r smonths$q50` months (IQR `r smonths$iqr` months). Each hospital contributed a median of `r pts_q$q50` patients (IQR `r pts_q$iqr`) equivalent to `r pts_overnight$q50` patients referred to critical care (IQR `r pts_overnight$iqr`) per 1,000 overnight admissions. Critical Care Outreach Teams operated 24 hours/day and 7 days/week in `r ccot$n[4]` (`r ccot$p[4]`) hospitals, less than 24 hours/day in `r ccot$n[3]` (`r ccot$p[3]`) hospitals, and less than 7 days/week in `r ccot$n[2]` (`r ccot$p[2]`) hospitals.

There was a median of `r beds$q50` (IQR `r beds$iqr`) critical care beds per hospital (mixed Level 2 --- typically intensive monitoring or single organ support, and Level 3 --- ventilated or multiple organ support), most often in a single physical location (`r colocate$n[1]` hospitals). 

## Incidence of critical care referrals
<!-- numbers here from stata log files for table -->

We estimated the mean incidence of referrals to critical care at 46 (95\%CI 50--54) patients per month of which 17 (95\%CI 17--18) patients would meet the definition of NEWS high risk at assessment. This was for a non-teaching hospital with 50,000 overnight admissions per year, and 24/7 CCOT provision. As critical care outreach provision decreased then the number of patients assessed fell (supplementary Table 1). Winter was busier (IRR 1.22, 95\%CI 1.14--1.31), and weekends quieter (IRR 0.87, 95\%CI 0.82-0.92) than non-winter weekday periods. When we included a measure of case finding in the models (cases assessed per 1000 overnight hospital admissions), referral incidence increased initially but may have begun to plateau for the hospitals with referral rates in the highest quartile (supplementary Figure 3).

## Patient characteristics
```{r include=FALSE, echo=FALSE}
source("5_results_03patients.R")
```

Table 1 shows the baseline data for all ward patients assessed. Sepsis was reported in (`r sepsis$n[2]`, `r sepsis$p[2]`) patients. Of these, the respiratory system was considered to be the source in half (`r s.resp.n`, `r s.resp.p`). Organ failure, defined as a SOFA score greater than or equal to two, was present in `r odys$n[2]` (`r odys$p[2]`) of patients. `r rdys$n[2]` patients (`r rdys$p[2]`) were in respiratory failure, `r kdys$n[2]` (`r kdys$p[2]`) were in renal failure, and `r shock$n[2]` (`r shock$p[2]`) were shocked. Organ support at the time of assessment was, however, uncommon (`r osupp$n[2]` patients, `r osupp$p[2]`).

Overall, `r dead7$n[2]` (`r dead7$p[2]`) patients died during the 7-days following ward assessment. Mortality was heavily front-loaded with `r dead7.d2$n[2]` (`r dead7.d2$p[2]`) of these deaths occuring in the first 48 hours (supplementary Figure 1). There was a clear correlation between physiological severity and early (7-day) mortality using either ward based (NEWS) or critical care scoring systems (SOFA, ICNARC) (supplementary Figure 2). For example, there were `r d7.news1$n[2]` (`r d7.news1$p[2]`), `r d7.news2$n[2]` (`r d7.news2$p[2]`), and `r d7.news3$n[2]` (`r d7.news3$p[2]`) deaths within a week for NEWS low, medium and high risk classes respectively.

## Patient pathways following ward assessment
```{r include=FALSE, echo=FALSE}
source("5_results_06pathways.R")
```

We extracted two clinical decisions at the point of the bedside assessment. We asked whether the bedside assessor would recommend the patient for critical care (`r recommend$n[2]` patients [`r recommend$p[2]`] recommended versus `r recommend$n[1]` patients [`r recommend$p[1]`] recommended for ward care). Additionally, we recorded the decision to admit, and classified patients as either initially refused (`r decision$n[2]` patients, `r decision$p[2]`), initially refused with treatment limitations (`r decision$n[3]` patients, `r decision$p[3]`), or immediately accepted to critical care (`r decision$n[1]` patients, `r decision$p[1]`). Amongst those recommended for critical care, the proportions initially refused (without treatment limits), initially refused with limits, or immediately accepted were `r decision.rec$p[2]` (`r decision.rec$n[2]` patients), `r decision.rec$p[3]` (`r decision.rec$n[3]` patients), and `r decision.rec$p[1]` (`r decision.rec$n[1]` patients) respectively.


We then tracked patient pathways over the following week.

### Patients initially refused critical care (without treatment limits)

Patients referred to, but then refused, critical care (without treatment limitations) had a 7-day mortality of `r ward.dead7$p[2]` (`r ward.dead7$n[2]` deaths). Most deaths (`r ward.dead7.icu$n[1]` deaths `r ward.dead7.icu$p[1]`)  occured on the ward without critical care but `r ward.dead7.icu$n[2]` deaths (`r ward.dead7.icu$p[2]`) followed delayed admission. Overall, the initial decision to refuse was reversed in `r ward.icucmp$n[2]` patients (`r ward.icucmp$p[2]`) so by the end of the week `r ward.icu.or.dead7$n[2]` (`r ward.icu.or.dead7$p[2]`) patients had died or been admitted to critical care.

7-day mortality was higher for those recommended but refused (`r ward.surv.by.reco$est2` versus `r ward.surv.by.reco$est1`, risk difference `r ward.surv.by.reco$d` 95\%CI `r ward.surv.by.reco$ci`), and for those where the initial decision to refuse was reversed (`r ward.surv.by.icu$est2` versus `r ward.surv.by.icu$est1`, risk difference `r ward.surv.by.icu$d` 95\%CI `r ward.surv.by.icu$ci`).

### Patients with treatment limits (and refused critical care)

These `r decision$n[3]` patients had a 7-day mortality of `r limits.dead7$p[2]` (`r limits.dead7$n[2]` deaths). The initial decision to refuse admission was reversed in just `r limits.icucmp$n[2]` patients (`r limits.icucmp$p[2]`) of whom `r limits.icu.dead7$n[2]` (`r limits.icu.dead7$p[2]`) died within the week.

Compared to those refused without a treatment limitation order, this cohort was older (`r age.by.limits$est2` vs `r age.by.limits$est1` years, 95\%CI for difference `r age.by.limits$ci` years), and more acutely unwell (`r sofa.by.limits$est2` vs `r sofa.by.limits$est1` SOFA points, 95\%CI for difference `r sofa.by.limits$ci`). The final 90-day mortality was substantial `r limits.dead90$p[2]` (`r limits.dead90$n[2]` deaths), nonetheless `r limits.dead1y$n[1]` patients (`r limits.dead1y$p[1]`) survived a year despite the decision.

### Patients immediately accepted to critical care

The `r decision$n[1]` (`r decision$p[1]`) patients accepted to critical care at the initial bedside assessment had a 7-day mortality of `r icu.dead7$p[2]` (`r icu.dead7$n[2]` deaths). Just `r icu.dead7.pre$n[1]` (`r icu.dead7.pre$p[1]`) of those deaths occurred before admission was arranged, but a further `r icu.icucmp$n[1]` patients (`r icu.icucmp$p[1]`) were never admitted but survived regardless.

Those offered admission were marginally younger (`r age.by.icu$est1` versus `r age.by.icu$est2` years, 95\%CI `r age.by.icu$ci` years), and more acutely unwell (`r sofa.by.icu$est1` versus `r sofa.by.icu$est2` SOFA points, 95\%CI `r sofa.by.icu$ci`).

The median delay between assessment and admission for these patients was `r icu.delay$q50` hours (IQR `r icu.delay$iqr`) compared to `r ward.delay$q50` hours (IQR `r ward.delay$iqr`) for those refused without treatment limits (median additional delay `r delay.by.accept$estimate` hours, IQR `r delay.by.accept$lower`--`r delay.by.accept$upper`). This meant that `r icu.early4$n[2]` (`r icu.early4$p[2]`) admissions were achieved within 4 hours for those initially accepted versus `r ward.early4$n[2]` (`r ward.early4$p[2]`) for those initially refused (risk difference `r early4.by.accept$d`, 95\%CI `r early4.by.accept$ci`). 


## Critical care occupancy
```{r include=FALSE, echo=FALSE}
source("5_results_05occupancy.R")
```

There were `r room$n[1]` (`r room$p[1]`) patients assessed when the unit was full, `r room$n[2]` (`r room$p[2]`) assessed when there were either one or two beds available, and `r room$n[3]` (`r room$p[3]`) assessed when there were more than two beds available (Table 2). Critical care occupancy fluctuated with time of the day, day of the week, and season of the year (supplementary Figure 2). 

As occupancy at the time of assessment increased, patients were less likely to be immediately, or ever, accepted to critical care; less likely to be admitted promptly; and, more likely to die on the ward without critical admission (Table 3, Cochran-Armitage test for trend p<0.0001 for all measures).

<!-- 
     _ __ ___  ___ _   _ _ __ ___   ___ 
    | '__/ _ \/ __| | | | '_ ` _ \ / _ \
    | | |  __/\__ \ |_| | | | | | |  __/
    |_|  \___||___/\__,_|_| |_| |_|\___|
 -->

### Variation in decision making

We built a multi-level (patients nested within hospitals) logistic regression model to examine factors associated with a decision to admit to critical care amongst those patients without a treatment limitation order. As with the univariate comparisons above, older patients were less likely to be admitted (OR 0.60 for patients over 80 years, 95\%CI 0.53--0.69), and sicker patients were more likely to be admitted (OR 1.07 per ICNARC physiology point, 95\%CI 1.06--1.07). Similarly, patients already receiving organ support (OR 95\%CI), or
clinically judged to be peri-arrest (OR 95\%CI) were much more likely to be admitted.

Patients referred overnight (7pm-7am), during the weekend, or during the winter were more likely to be offered critical care, but those assessed when the critical care unit was at, or near capacity were less likely to be admitted (OR 0.70, 95\%CI 0.57--0.86, and OR 0.87 95\%CI 0.77--0.98 respectively). Otherwise said, had there been no limitations on capacity then an additional 122 patients (95\%CI 53-186) would have been admitted

The model also demonstrated significant hospital level variation with a Median Odds Ratio (MOR) between hospitals of 1.82 (95\%CI 1.78-2.16) which was minimally affected by adjustment for patient characteristics (MHR 1.82 95\%CI 1.78--2.16 in a model excluding these). This implies that when comparing patients from any two randomly selected hospitals the adjusted odds of being offered critical care varied by almost two fold.

Mortality
---------

There were 706 deaths by the end of the first day, 2,787 by the end of the first week, 4,561 by the end of the first month, and 6,989 by the end of the first year. The risk of death at these time points was 4·5\%, 17·9\%, 29·2\%, and 44·8\% respectively (Kaplan-Meier failure function). The period of greatest risk immediately followed the referral, falls rapidly, but remained elevated even at one-year (excess mortality 0·103 deaths/patient-year, 95\%CI 0·076--0·130.([Figure 4][figure4]).[Cancer Research UK Cancer Survival Group, 2006, \#91622]

During the week following the ward visit, the majority (1767, 63·4\%) of deaths occurred without admission to critical care. The majority of these patients (934, 52·9\%) had no treatment limitation order reported at the time of their initial evaluation ([Figure 5][figure5]). These patients were less unwell (SOFA score -0·68, 95\%CI -0·46 to -0·91), but older (6·3 years, 95\%CI 5·1--7·5) than those deaths occurring in ICU ([supplementary Table 4](<stb04_ward_vs_icu_deaths.ods>)). A small proportion (66, 7·1\%) of the ward deaths had been accepted to critical care, but died before the admission was realised. In contrast, among those with treatment limitation orders, 472 (23·1\%) patients survived to one year without admission to critical care in that first week.

A series of models were fitted with 90-day survival as the dependent variable. The final best model [Table 4](<tb03_ward_survival_final.ods>) incorporated a time-varying effect for acute physiological severity so that the effect of severity on survival was greatest in the first days following assessment (see supplementary [supplementary Figure 1](<sfig01_survival_icnarc0_ssresidual.pdf>)). Other patient level risk factors were consistent with the existing literature on outcomes in similar patients: older patients, males, patients receiving a higher level of care, and those with sepsis (other than genito-urinary) had worse outcomes.[Harrison, 2007] Patients admitted during the winter months had an adjusted hazard ratio of 1·12 (95\%CI 1·04--1·20), but neither time of day, nor day of the week affected survival. 

At the hospital level, with the exception of the two hospitals without a CCOT service, there was weak stepwise decrease in survival as the provision of CCOT services expanded. No other hospital level factors were associated with outcome. The final best model demonstrated significant hospital level variation (variance 0·030, 95\%CI 0·013--0·046, [supplementary Figure 2](<sfig02_survival_reffects.pdf>)). This gives a Median Hazard Ratio (MHR) between hospitals of 1·28 in the final model which was not markedly affected by adjustment for patient characteristics (MHR 1·29 in a model excluding these).

