---
title: "5_results"
author: "Steve Harris"
date:	"30 Dec 2015"
output: md_document
---
```{r setup, include=FALSE, echo=FALSE}
source("5_results_01start.R")
```

# Results

48 hospitals reported 20893 visits over 435 months. 2694 visits (12.9\%) did not meet the inclusion criteria including 1860 (8.9\%) repeat visits, and 586 (2.8\%) patients recently discharged from critical care. Data linkage was incomplete (\< 80\%) for 66 (15\%) study-months excluding a further 2440 (11.7\%) visits. Therefore 15759 patients were recruited to the study, of which 15158 (96.1\%) completed follow-up without error and were available for analysis (Figure 1a). Final data linkage (ward visits to critical care admissions) was 93\% complete.

## Participating hospitals
```{r hospitals, include=FALSE, echo=FALSE}
source("5_results_04hospitals.R")
```
```{r occupancy, include=FALSE, echo=FALSE}
source("5_results_05occupancy.R")
```

The participating hospitals comprised `r n.teach` teaching and `r n.dgh` general hospitals that each collected data between September 2010 and December 2011 for a median of `r smonths$q50` months (IQR `r smonths$iqr` months). Each hospital contributed a median of `r pts_q$q50` patients (IQR `r pts_q$iqr`) equivalent to `r pts_overnight$q50` patients referred to critical care (IQR `r pts_overnight$iqr`) per 1,000 overnight admissions. Critical Care Outreach Teams (CCOT) operated 24 hours/day and 7 days/week in `r ccot$n[4]` (`r ccot$p[4]`) hospitals, less than 24 hours/day in `r ccot$n[3]` (`r ccot$p[3]`) hospitals, and less than 7 days/week in `r ccot$n[2]` (`r ccot$p[2]`) hospitals. Two hospitals had no CCOT.

There was a median of `r beds$q50` (IQR `r beds$iqr`) critical care beds per hospital (mixed Level 2 [typically intensive monitoring or single organ support], and Level 3 [ventilated or multiple organ support]), most often colocated in a single physical location (`r colocate$n[1]` hospitals).

The critical care unit was fully occupied for `r room$n[1]` (`r room$p[1]`) assessments, under strain for  `r room$n[2]` (`r room$p[2]`) assessments (one or two beds available), and had three or more beds available for the remaining `r room$n[3]` (`r room$p[3]`) assessments. Critical care occupancy fluctuated with time of the day, day of the week, and season of the year (supplementary Figure 1).

## Incidence of critical care referrals
<! to  numbers here from stata log files for incidence models  to >

We estimated the mean baseline (non-teaching hospital with 50,000 admissions per year and 24/7 CCOT provision) incidence of referrals to critical care at 46 (95\% confidence interval 50 to 54) patients per month of which 17 (95\%CI 17 to 18) patients met the NEWS high risk criteria at assessment. As critical care outreach provision decreased then the number of patients assessed fell (supplementary Table 1). Winter was busier (IRR 1.22, 95\%CI 1.14 to 1.31), and weekends quieter (IRR 0.87, 95\%CI 0.82 to 0.92) than the rest of year. When we included a measure of case finding in the models (cases assessed per 1000 overnight hospital admissions), referral incidence increased initially but may have begun to plateau for the hospitals with referral rates in the highest quartile (supplementary Figure 2).

## Patient characteristics
```{r patients, include=FALSE, echo=FALSE}
source("5_results_03patients.R")
```

Table 1 shows the baseline data for all ward patients assessed. Sepsis was reported in `r sepsis$n[2]` patients (`r sepsis$p[2]`). Of these, the respiratory system was considered to be the source in half (`r s.resp.n`, `r s.resp.p`). Organ failure, defined as a SOFA score greater than or equal to two, was present in `r odys$n[2]` of patients (`r odys$p[2]`). `r rdys$n[2]` patients (`r rdys$p[2]`) were in respiratory failure, `r kdys$n[2]` (`r kdys$p[2]`) were in renal failure, and `r shock$n[2]` (`r shock$p[2]`) were shocked. Organ support at the time of assessment was uncommon (`r osupp$n[2]` patients, `r osupp$p[2]`).

`r dead7$n[2]` (`r dead7$p[2]`) patients died during the 7-days following ward assessment. Mortality was heavily front-loaded with `r dead7.d2$n[2]` (`r dead7.d2$p[2]`) of these deaths occuring in the first 48 hours (supplementary Figure 3). There was a clear correlation between physiological severity and early (7-day) mortality using either ward based (NEWS) or critical care scoring systems (SOFA, ICNARC) (supplementary Figure 3). For example, the 7-day mortality was `r d7.news1$p[2]` (`r d7.news1$n[2]` deaths), `r d7.news2$p[2]` (`r d7.news2$n[2]` deaths), and `r d7.news3$p[2]` (`r d7.news3$n[2]` deaths)for NEWS low, medium and high risk classes respectively.

The critical care assessor judged that `r recommend$n[2]` patients (`r recommend$p[2]`) required critical care. These patients had a higher physiological severity of illness (ICNARC physiology score `r icnarc.by.reco$est2` versus `r icnarc.by.reco$est1`, 95% confidence interval for difference `r icnarc.by.reco$ci`), and a greater 7-day mortality `r d7.reco$est2` versus `r d7.reco$est1`, difference `r d7.reco$ci`. There was a clear correlation between the recommendation and measured severity (supplementary Figure 5).

Overall mortality at 90-days and 1-year was `r dead90$p[2]` (`r dead90$n[2]` patients), and `r dead1y$p[2]` (`r dead1y$n[2]` patients) respectively.

## Patient pathways following ward assessment
```{r pathways, include=FALSE, echo=FALSE}
source("5_results_06pathways.R")
```

Patients were classified into three groups following the ward assessment: those refused admission with treatment limits (pre-existing or otherwise); those refused admission without treatment limits (active ward care); and, those offered immediate critical care (Figure 1).

There were `r decision$n[3]` patients (`r decision$p[3]`) refused with treatment limits, `r decision$n[2]` patients (`r decision$p[2]`) assigned to active ward-care,  and `r decision$n[1]` patients (`r decision$p[1]`) immediately offered critical care.

### Refused critical care with treatment limits

The `r decision$n[3]` patients with treatment limits had a 7-day mortality of `r limits.dead7$p[2]` (`r limits.dead7$n[2]` deaths). The initial decision to refuse admission was reversed in just `r limits.icucmp$n[2]` patients (`r limits.icucmp$p[2]`) of whom `r limits.icu.dead7$n[2]` (`r limits.icu.dead7$p[2]`) died within the week. Although the final 90-day mortality was substantial `r limits.dead90$p[2]` (`r limits.dead90$n[2]` deaths), `r limits.dead1y$n[1]` patients (`r limits.dead1y$p[1]`) survived a year despite the decision.

These patients were older (`r age.by.limits$est2` versus `r age.by.limits$est1` years, 95\%CI for difference `r age.by.limits$ci` years), and more acutely unwell (`r icnarc.by.limits$est2` vs `r icnarc.by.limits$est1` ICNARC points, difference `r icnarc.by.limits$ci`) than those assigned active ward care. Critical care occupancy did not affect the proportion of patients refused admission with treatment limits (Table 2).<! to   We did not consider these patients further.  to >

### Active ward care

The `r decision$n[2]` patients for active ward care had a 7-day mortality of `r ward.dead7$p[2]` (`r ward.dead7$n[2]` deaths). Most deaths (`r ward.dead7.icu$n[1]` deaths `r ward.dead7.icu$p[1]`)  occured on the ward but `r ward.dead7.icu$n[2]` deaths (`r ward.dead7.icu$p[2]`) followed delayed critical care admission. The initial decision to refuse critical care was reversed for `r ward.icucmp$n[2]` patients (`r ward.icucmp$p[2]`), so a total of `r ward.icu.or.dead7$n[2]` (`r ward.icu.or.dead7$p[2]`) patients died or were admitted to critical care.

The active ward care group included `r ward.reco$n[2]` (`r ward.reco$p[2]`) patients who had been recommended critical care on assessment. These patients had a higher 7-day mortality (`r ward.surv.by.reco$est2` versus `r ward.surv.by.reco$est1`,  95% confidence interval for difference `r ward.surv.by.reco$ci`), and were more likely to have the initial refusal reversed (`r ward.icu.by.reco$est2` versus `r ward.icu.by.reco$est1`,  risk difference `r ward.icu.by.reco$ci`). 

### Immediate critical care

The `r decision$n[1]` patients immediately accepted to critical care had a 7-day mortality of `r icu.dead7$p[2]` (`r icu.dead7$n[2]` deaths). Just `r icu.dead7.pre$n[1]` (`r icu.dead7.pre$p[1]`) of those deaths occurred before admission was arranged, but a further `r icu.icucmp$n[1]` patients (`r icu.icucmp$p[1]`) were never admitted but survived nonetheless.

Those offered immediate admission were marginally younger (`r age.by.icu$est1` versus `r age.by.icu$est2` years, 95\%CI `r age.by.icu$ci` years), but distinctly more unwell (`r icnarc.by.icu$est1` versus `r icnarc.by.icu$est2` ICNARC physiology points, 95\%CI `r icnarc.by.icu$ci`) than the active ward care group. As critical care occupancy at the time of the ward assessment increased, patients were less likely to be immediately accepted (Table 2, Cochran-Armitage test for trend p<0.0001).

## Delay to admission to critical care
<! to  resume  to >

The median delay between assessment and admission for patients immediately accepted was `r icu.delay$q50` hours (IQR `r icu.delay$iqr`) compared to `r ward.delay$q50` hours (IQR `r ward.delay$iqr`) for late admissions from the active ward care group (median additional delay `r delay.by.accept$estimate` hours, IQR `r delay.by.accept$lower` to `r delay.by.accept$upper`) (Figure 2a). This meant that `r icu.early4$n[2]` (`r icu.early4$p[2]`) admissions were achieved within 4 hours for those immediately accepted versus `r ward.early4$n[2]` (`r ward.early4$p[2]`) for those initially refused (risk difference `r early4.by.accept$d`, 95\%CI `r early4.by.accept$ci`). For the subgroup recommended to critical care whose initial refusal was later reversed, the median delay to admission was `r ward.delay.reco$q50` hours (IQR `r ward.delay.reco$iqr`).

Increasing occupancy at assessment increased the median delay from 3, to 4, to 6 hours as available beds fell from more than three to none (Figure 2b, Cochran-Armitage test for trend p<0.0001).

## Determinants of a decision to admit
```{r decision, include=FALSE, echo=FALSE}
source("5_results_07models.R")
```

We built a multi-level (patients nested within hospitals) logistic regression model to examine factors associated with a decision to admit for patients without treatment limits (Table 3). As with the univariate comparisons above, older patients were less likely to be admitted (patients over 80 years: odds ratio 0.60, 0.53 to 0.69), and more acutely unwell patients were more likely to be admitted (OR 1.07 per ICNARC physiology point (e.g. an increase in heart rate from 105 to 115), 95\%CI 1.06 to 1.07). Similarly, patients already receiving organ support (1.83, 1.55 to 2.16), or clinically judged to be peri-arrest (6.32, 5.18 to 7.70) were also more likely to be admitted.

Patients referred out-of-hours (7pm-7am), during the weekend, or during the winter were more likely to be offered critical care (odds ratios between 1.04 to 1.33), but those assessed when the critical care unit was full (0.70,  0.57 to 0.86), or near full were less likely to be admitted (0.87, 0.77 to 0.98). We estimated that in this sample had there been no limitations on capacity then an additional 122 patients (95\%CI 53 to 186) would have been admitted.

The model also demonstrated significant hospital level variation with a Median Odds Ratio (MOR) of 2.11 (95\%CI 1.81 to 2.42) which differed little to that estimated excluding patient predictors (MOR 2.18, 1.82 to 2.60). The MOR summarises the differences when comparing decision making for patients from any two randomly selected hospitals, and consistency when excluding patient level predictors suggests that it is a true hospital level difference.

## Determinants of prompt admission

The modelling was repeated but now with the delivery of admission to critical care within 4 hours (a prompt admission) as the outcome, and the decision to admit as an additional predictor (Table 3). We excluded an additional `r elgthtr$n[3]` (`r elgthtr$p[3]`) patients where surgery between assessment and admission inevitably delayed that admission.

The associations between prompt admission and patient level predictors were broadly similar with younger and sicker patients being admitted more promptly. However, patients assessed during the winter, while being more likely to be offered critical care, were now less likely to be admitted promptly (odds ratio 0.76, 0.64 to 0.90). The strongest predictor of prompt admission was a decision to admit at the initial bedside assessment (69, 59 to 81). Even though the decision to admit was included in the model, occupancy still had a marked effect (zero beds: 0.27, 0.19 to 0.37) and hospital level variation persisted (MOR 1.89, 1.63 to 2.21).

## Determinants of 90-day mortality
```{r survival, include=FALSE, echo=FALSE}
source("5_results_08survival.R")
```

Amongst patients without treatment limits, there were `r t.surv$n[2]` deaths by the end of the first day, `r t.surv$n[3]` by the end of the first week, `r t.surv$n[4]` by the end of the first month, and `r t.surv$n[5]`  by 90 days. The risk of death at these time points was `r t.surv$fail[2]`, `r t.surv$fail[3]`, `r t.surv$fail[4]`, and `r t.surv$fail[5]` respectively (Kaplan-Meier failure function).

A series of models were fitted with 90-day survival as the dependent variable for patients without treatment limits. The final best model (Table 4) incorporated a time-varying effect for acute physiological severity such that its effect was attenuated after the first week (supplementary Figure 6).

Other patient level risk factors were consistent with the existing literature on outcomes in similar patients: older patients, and those with sepsis (other than genito-urinary) had worse outcomes.[@Harrison:2004brb] Patients assessed during the winter months, over the weekend, and out-of-hours did not have a significantly worse adjusted survival than baseline.

Critical care occupancy did not affect adjusted mortality in the multi-level model (zero beds: hazard ratio 1.03, 95\% confidence interval 0.90 to 1.17). A single level model (supplementary Table 1), constructed in case occupancy was mediated through rather than confounded by the effect of the hospital, similarly could not exclude a null effect (1.07, 1.00 to 1.15).

The full multi-level model demonstrated significant hospital level variation in survival (median hazard ratio 1.28, 1.22 to 1.34) which was little altered by adjustment for patient level risk factors (MHR 1.29, 1.22 to 1.35).

Survival models in the subgroup recommended to critical care at the initial assessment similar effects albeit with less precision (supplementary Table 2).
